Emergence of the Southeast Asian islands as a driver for Neogene cooling

### 2020-04-30 Manuscript/Manuscript.tex
Updates from Overleaf

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{Author contributions: Y.P., P.M., Y.G., and N.L.S-H. designed and implemented the GEOCLIM model. F.A.M. and E.A.A. constructed the paleoshoreline database. Y.P. and P.M. executed the model and analyzed the data. Y.P., P.M., Y.G., F.A.M., and N.L.S-H. wrote the manuscript.}
\authordeclaration{The authors declare no competing interest.}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: yuempark@berkeley.edu}

@@

Topography, climate, and lithology all effect chemical weathering. High-relief regions lack extensive regolith development, and thus tend to have reaction-limited weathering regimes that are more sensitive to climate change \cite{Gabet2009a, West2012a, Maher2014a}. The high physical erosion rates contribute to high chemical weathering fluxes in these high-relief regions \citep{Godderis2017b}. In warm and wet regions, mineral dissolution kinetics are faster leading to enhanced chemical weathering \cite{Lasaga1994a, West2012a}. Mafic rocks have higher Ca and Mg concentrations and dissolution rates than felsic rocks, and thus have the potential to more efficiently sequester carbon through silicate weathering \cite{Dessert2003a}. These factors have lead to the proposal that arc-continent collisions, which create steep landscapes of mafic lithologies, within the tropical rain belt have been important in enhancing global weatherability, lowering atmospheric \pCOtwo, and initiating glacial climate over the past 520~m.y. \cite{Jagoutz2016a, Swanson-Hysell2017a, Macdonald2019a} and perhaps in the Neoproterozoic as well \cite{Park2019b}.

Quantitatively estimating the magnitude of decrease in steady-state \pCOtwo associated with the emergence of a region with a high carbon sequestration potential, such as the SEAI, is not straight-forward. As this region emerges, the total global silicate weathering flux will briefly exceed the volcanic degassing flux, causing \pCOtwo to initially decline. However, the sensitivity of the silicate weathering flux in any particular location to this change in \pCOtwo is highly variable and dependent on the specific topography, climate, and lithology at that location. Furthermore, how the local climate responds to this change in \pCOtwo is itself highly spatially-variable. Therefore, the magnitude of \pCOtwo change that is required to balance the total global silicate weathering flux with the volcanic degassing flux will depend on the specific spatial distribution of topography, climate, and lithology at the time of emergence. Ultimately, there is a complex, non-linear, and spatially-varying system of feedbacks between climate and silicate weathering. Therefore, any attempt to meaningfully estimate the decrease in steady-state \pCOtwo associated with emergence of the SEAI must model spatially-resolved climatology and silicate weathering fluxes in tandem and account for the spatial distribution of the factors that affect these two systems.

@@

To estimate the decrease in steady-state \pCOtwo associated with the increase of subaerially-exposed land area in the SEAI, we use the globally spatially-resolved GEOCLIM model \cite{Godderis2017c}. GEOCLIM estimates changes in steady-state \pCOtwo associated with coupled changes in erosion, chemical weathering, and climatology by linking a silicate weathering model to climate model runs at multiple \pCOtwo levels.

@@

The silicate weathering component of GEOCLIM calculates \COtwo consumption resulting from silicate weathering for subaerially-exposed land. We assume that Ca and Mg are the only cations that consume \COtwo over geologic time-scales, such that each mole of Ca or Mg that is dissolved by silicate weathering consumes one mole of \COtwo. In previous versions of the model, silicate weathering was a function of temperature and runoff only, and all bedrock was assigned identical chemical compositions \cite{Godderis2017c}. More recent versions of GEOCLIM implement regolith development and soil shielding (Fig. \ref{fig:regolith_schematic}), which introduces a dependence on erosion (and therefore topographic slope) \cite{Maffre2018a}. While this introduction of regolith development into GEOCLIM is important for assessing the impact of tropical arc-continent collisions on \pCOtwo, the relatively high Ca+Mg concentration in arc rocks relative to other lithologies must also be considered.

@@

The values of four parameters within the silicate weathering component that modify the dependence of silicate weathering on temperature, runoff, and regolith thickness are poorly constrained. Rather than prescribing values, we allow these parameters and the Ca+Mg concentration of metamorphic and siliciclastic pixels to vary within reasonable bounds (\SI; Table S2). This results in 93,600 unique combinations of these six parameters. For each combination, we compute spatially-resolved long-term \COtwo consumption associated with Ca+Mg fluxes using present-day runoff, temperature, and slope. We sum computed \COtwo consumption over watersheds for which data-constrained estimates are available \cite{Gaillardet1999a, Moquet2018a}, then calculate the coefficient of determination ($r^{2}$) between computed and measured \COtwo consumption in each of these watersheds. After eliminating parameter combinations that result in low $r^{2}$, 573 parameter combinations remain (\SI; Fig. S2). The resulting global \COtwo consumption of these model runs all overlap with independently-derived estimates of the global \COtwo degassing flux \cite{Gerlach2011a} as they should given that the long-term carbon cycle is in steady state (\SI; Fig. S2).

@@

Having calibrated the silicate weathering component of GEOCLIM, we use it to estimate the decrease in steady-state \pCOtwo associated with SEAI emergence. For the climate model component, we use temperature and runoff from a subset of the GFDL CM2.0 experiments \cite{Delworth2006b} (\SI). These experiments are useful for this analysis because all non-\COtwo forcings are held constant at values representative of pre-Industrial conditions, allowing the effect of changing \pCOtwo on climatology to be isolated. Furthermore, the experiments were run long enough for the final system to approximate steady-state.

@@

\begin{SCfigure*}[\sidecaptionrelwidth][t]
    \centering
    \includegraphics[width=13.0cm]{Figures/shoreline_growth.pdf}
    \caption{The emergence of the Southeast Asian islands (also referred to as the Maritime Continent in climate science literature) from the mid-Miocene to present. Past shorelines are shown in \textit{A} with associated land area summarized in \textit{B}. A significant increase in area over the past 5 million years is coincident with cooling and the onset of Northern Hemisphere glaciation as reflected in the benthic oxygen isotope record \cite{Zachos2008a} shown in \textit{C}.}
    \label{fig:shoreline_growth}
\end{SCfigure*}

To determine the position of SEAI paleoshorelines over the past 15~m.y., we use terrestrial and marine sedimentary deposits (Fig. \ref{fig:shoreline_growth}; \SI). The paleoshoreline data indicate that the Sunda-Banda Arc and New Guinea are primarily responsible for the increase in area since 15~Ma. Exhumation of the modern Sunda-Banda Arc is the result of ongoing arc-continent collision with the subducting Australian Plate \cite{Harris2006a}. Of the major islands in this arc, only Sumatra, Belitung, Bangka, Java, Bali, and Flores were subaerially exposed before 5~Ma. However, most of Sumatra and Java along with the non-volcanic islands of the Outer Banda Arc were elevated above sea level after 5~Ma \cite{Hall2013b}. In New Guinea, emergence in the mid-Miocene is associated with collision between the Melanesian Arc and Australia's distal margin \cite{Cloos2005a}, which drove exhumation of the Irian-Marum-April Ophiolite Belt. Exhumation accelerated over the past 4~m.y. in the New Guinea Central Range due to slab-breakoff and buoyant uplift, and in eastern New Guinea due to jamming of the north-dipping subduction zone \cite{Cloos2005a}. These tectonic drivers and others throughout the region lead to progressive emergence over the past 15~m.y. that accelerated following 5~Ma (Fig. \ref{fig:shoreline_growth}B). This trend mirrors broad cooling over the Neogene that resulted in the initiation of Northern Hemisphere ice sheets (Fig. \ref{fig:shoreline_growth}C). We also include changes in areas of presently-submerged continental shelves such as the Sunda Shelf that were previously exposed (\SI; Fig. S5).

@@

Proxy-based estimates of the magnitude and trajectory of \pCOtwo change from the Miocene to the Pliocene are variable between techniques and associated assumptions underlying their interpretation \cite{Foster2017a}. The modeled \pCOtwo values for 15~Ma resemble the higher end of proxy-based \pCOtwo estimates for the early-mid-Miocene (\SI; Fig. S9), indicating that the increase in subaerially-exposed land area and tectonic topography of the SEAI is sufficient to explain long-term cooling of Earth's climate over the Neogene. The \pCOtwo threshold for Antarctic glaciation is estimated to be $\sim$750~ppm with that for Northern Hemisphere glaciation being significantly lower at $\sim$280~ppm \cite{DeConto2008a}. These modeled values of decreasing \pCOtwo are therefore consistent with the record of Neogene climate with Miocene ice sheets on Antarctica \cite{Sugden1995a} followed by Northern Hemisphere ice sheets developing in the Pliocene \cite{Haug2005a} as \pCOtwo subsequently decreased.

@@

Previous work has estimated that the decrease in \pCOtwo since 5~Ma associated with the emergence of the SEAI and enhanced silicate weathering is $\sim$19~ppm \cite{Molnar2015a}, in which case SEAI emergence would be a relatively minor contributor to Neogene cooling. This 19~ppm estimate was obtained using a 0-D equation that assumes a direct linear relationship between mean global temperature and changes in weathering-rate-weighted land area, scaled by a factor that is intended to account for the influence of both runoff and temperature. \pCOtwo was then estimated from the calculated temperature using a simple energy balance equation. However, the relationship between mean global temperature (or \pCOtwo) and weathering-rate-weighted land area is not linear. Furthermore, this simple linear relationship ignores spatial variability in topography and climatology, and only crudely accounts for spatial variability in lithology. In fact, the 19~ppm estimate is closer in magnitude to the decrease in \pCOtwo that we estimate if mafic volcanics associated with the Deccan Traps (a relatively flat area outside of the warm and wet tropics) are replaced with the Ca+Mg concentration of bulk continental crust (22--70 ppm; Fig. \ref{fig:scenario_pCO2}). The significant difference in steady-state \pCOtwo estimated between the ``removed Deccan Traps'' scenario and the ``paleo-SEAI'' scenarios (Fig. \ref{fig:scenario_pCO2}) demonstrates that considering changes in the spatial distribution of lithologies alone is not adequate for estimating changes in steady-state \pCOtwo. Instead, spatially-varying topography and climatology significantly modulates silicate weathering rates, and must be accounted for.

\section*{Alternative Mechanisms}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=1\linewidth]{Manuscript/Figures/weatherability_curves.pdf}
    \caption{Weatherability curves for the ``paleo-SEAI'' scenarios shown in Figure \ref{fig:scenario_pCO2}. Details on how these curves were generated are described in \MM. Each of the 4 curves represent a different tectonic boundary condition and therefore a different global weatherability (i.e. the reconstructed paleoshorelines of the SEAI; Fig. \ref{fig:shoreline_growth}A), and shows the volcanic degassing flux necessary to balance the silicate weathering flux at a given \pCOtwo and global weatherability. Point B represents the pre-Industrial, in which \pCOtwo is 286~ppm. The arrow from Point A$_{1}$ to B represents the ``change in weatherability only'' scenario, in which global weatherability changes as the SEAI emerge, but the volcanic degassing flux does not change over the past 15~m.y. In this scenario, the \pCOtwo decreases from the value dictated by the 15~Ma ``paleo-SEAI'' weatherability curve ($\sim$570~ppm). The arrow from Point A$_{2}$ to B instead represents the ``change in degassing only'' scenario, in which global weatherability remains the same as the pre-Industrial, but the same change in \pCOtwo as the ``change in weatherability only'' scenario is achieved by decreasing the volcanic degassing flux from a value $\sim$13\% greater than the pre-Industrial. The arrow from Point A$_{3}$ to B represents the ``change in weatherability and degassing'' scenario, in which the same change in \pCOtwo as the other two scenarios is achieved by increasing global weatherability from our 5~Ma tectonic boundary condition and decreasing the volcanic degassing flux from a value $\sim$7\% greater than the pre-Industrial.}
    \label{fig:weatherability_curves}
\end{figure}

@@

Nevertheless, our model framework provides us with the opportunity to estimate the decrease in volcanic degassing flux necessary to achieve the same change in \pCOtwo predicted for the increase in global weatherability associated with the emergence of the SEAI over the past 15~m.y. If we use the parameter combination that had the highest $r^{2}$ between computed and measured \COtwo consumption in watersheds around the world during calibration (\textit{Calibration} and \SI; Fig. S3), GEOCLIM estimates a volcanic degassing flux for the pre-Industrial of 4.1$\times$10$^{12}$~mol/yr. If we then assume that this volcanic degassing flux did not change over the past 15~m.y., then GEOCLIM estimates that the increase in global weatherability associated with the emergence of the SEAI lead to a change in \pCOtwo of $\sim$280~ppm (``change in weatherability only'' scenario in Fig. \ref{fig:weatherability_curves}). If we instead assume that global weatherability did not change over the past 15~m.y., then we estimate that the volcanic degassing flux needs to have been $\sim$13\% greater at 15~Ma relative to the pre-Industrial to drive the same $\sim$280~ppm change in \pCOtwo (``change in degassing only'' scenario in Fig. \ref{fig:weatherability_curves}).

This $\sim$13\% value is higher than 10\%, the highest current estimate for the volcanic degassing flux at 15~Ma relative to the present day \cite{Van-Der-Meer2014a}, suggesting that reasonable changes in the volcanic degassing flux alone are insufficient to drive the majority of cooling observed over the Neogene. However, changes in the volcanic degassing flux would have modulated changes in \pCOtwo associated with changes in global weatherability. For example, let us assume that we have overestimated the increase in global weatherability associated with SEAI emergence over the past 15~m.y., and instead explore a scenario in which this increase is more accurately reflected by the difference in global weatherability between our 5~Ma and pre-Industrial tectonic boundary conditions (``change in weatherability and degassing'' scenario in Fig. \ref{fig:weatherability_curves}). In this scenario, if the volcanic degassing flux decreased from a value $\sim$7\% greater than the pre-Industrial, we can achieve the same $\sim$280~ppm change in \pCOtwo.

@@

Himalayan uplift \cite{Raymo1992a, Galy2007a} has also been proposed as drivers for Neogene cooling. Rising marine \SrSr has been associated with increasing weathering of the high \SrSr Himalayan lithologies \cite{Raymo1992a}. This increase in \SrSr since ca. 35~Ma is followed by a decrease in slope ca. 15~Ma that has been attributed to exhumation of relatively low \SrSr and high \OsOs Lesser Himalayan lithologies \cite{Quade1997a, Colleps2018a}, but could also be at least partially driven by the exhumation of mafic and organic-rich forearc sediments in the SEAI during arc-continent collision. Enhanced organic matter burial in the Bay of Bengal due to high sedimentation rates could also have contributed to Neogene cooling \cite{Galy2007a}.

@@

Coupled geological constraints and modeling experiments demonstrate that the SEAI has been a growing hot spot for carbon sequestration from the Miocene to present. Changes in volcanic degassing and paleogeography elsewhere on Earth, particularly in the Himalaya and Central America, would have also effected geological carbon sources and sinks. Such changes are likely to have induced complex feedbacks on many Earth surface processes. Yet, not only does the history of SEAI emergence coincide with Neogene cooling and the onset of Northern Hemisphere glaciation, but our coupled weathering-climate model also indicates that the associated steady-state \pCOtwo change is sufficient to explain the majority of this cooling. These results highlight that the Earth's climate state is particularly sensitive to changes in tropical paleogeography.

@@

To create the 15~Ma ``paleo-SEAI'' curve shown in Figure \ref{fig:weatherability_curves}, we use the reconstructed paleoshorelines of the SEAI at 15~Ma (Fig. \ref{fig:shoreline_growth}A). We then select the parameter combination that had the highest $r^{2}$ between computed and measured \COtwo consumption in watersheds around the world during calibration (\textit{Calibration} and \SI; Fig. S3), and fix \pCOtwo at the 3 \pCOtwo levels at which the GFDL CM2.0 climate model experiments were computed (\SI). We then run GEOCLIM at each of these \pCOtwo levels until steady state is achieved (i.e. until the volcanic degassing flux is equal to the silicate weathering flux). We then repeat this process for the 10 and 5~Ma ``paleo-SEAI'' paleoshorelines and the present day shorelines to generate the 3 other weatherability curves.

@@

\acknow{Collaborative research between NLS-H and YG was initially supported by a grant from the France-Berkeley Fund. FAM and NLS-H gratefully acknowledge support through NSF FRES grants \#1926001 and \#1925990. We thank Alec Brenner, Sam Lo Bianco, Mariana Lin, and Judy Pu for their data compilation contributions to the paleoshoreline reconstructions.}

>>>>

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{Author contributions: Y.P., P.M., Y.G., and N.L.S.-H. designed and implemented the GEOCLIM model. F.A.M. and E.A.A. constructed the paleoshoreline database. Y.P. and P.M. executed the model and analyzed the data in consultation with N.L.S.-H and Y.G.. Y.P., P.M., Y.G., F.A.M., and N.L.S.-H. wrote the manuscript.}
\authordeclaration{The authors declare no competing interest.}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: yuempark@berkeley.edu}

@@

Topography, climate, and lithology all effect chemical weathering. High-relief regions lack extensive regolith development, and thus tend to have reaction-limited weathering regimes that are more prone to adjust when climate changes \cite{Gabet2009a, West2012a, Maher2014a}. High physical erosion rates contribute to high chemical weathering fluxes in these high-relief regions \citep{Godderis2017b}. In warm and wet regions, mineral dissolution kinetics are faster leading to enhanced chemical weathering \cite{Lasaga1994a, West2012a}. Mafic rocks have higher Ca and Mg concentrations and dissolution rates than felsic rocks, and thus have the potential to more efficiently sequester carbon through silicate weathering \cite{Dessert2003a}. These factors have lead to the proposal that arc-continent collisions, which create steep landscapes of mafic lithologies, within the tropical rain belt have been important in enhancing global weatherability, lowering atmospheric \pCOtwo, and initiating glacial climate over the past 520~m.y. \cite{Jagoutz2016a, Swanson-Hysell2017a, Macdonald2019a} and perhaps in the Neoproterozoic as well \cite{Park2019b}.

Quantitatively estimating the magnitude of decrease in steady-state \pCOtwo associated with the emergence of a region with a high carbon sequestration potential, such as the SEAI, is not straight-forward. As this region emerges, the total global silicate weathering flux will briefly exceed the volcanic degassing flux, causing \pCOtwo to initially decline. However, the sensitivity of the silicate weathering flux in any particular location to this change in \pCOtwo is variable and dependent on the specific topography, climate, and lithology at that location. Furthermore, how the local climate responds to this change in \pCOtwo is itself spatially-variable. Therefore, the magnitude of \pCOtwo change that is required to balance the total global silicate weathering flux with the volcanic degassing flux will depend on the specific spatial distribution of topography, climate, and lithology at the time of emergence. As a result, any attempt to meaningfully estimate the decrease in steady-state \pCOtwo associated with emergence of the SEAI must model spatially-resolved climatology and silicate weathering fluxes in tandem and account for the spatial distribution of the factors that affect these inter-connected systems.

@@

To estimate the decrease in steady-state \pCOtwo associated with the increase of subaerially-exposed land area in the SEAI, we use the global spatially-resolved GEOCLIM model \cite{Godderis2017c}. GEOCLIM estimates changes in steady-state \pCOtwo associated with coupled changes in erosion, chemical weathering, and climatology by linking a silicate weathering model to climate model runs at multiple \pCOtwo levels.

@@

The silicate weathering component of GEOCLIM calculates \COtwo consumption resulting from silicate weathering for subaerially-exposed land. We assume that Ca and Mg are the only cations that consume \COtwo over geologic time-scales, such that each mole of Ca or Mg that is dissolved by silicate weathering consumes one mole of \COtwo. In previous versions of the model, silicate weathering was a function of temperature and runoff only, and all bedrock was assigned identical chemical compositions \cite{Godderis2017c}. More recent versions of GEOCLIM implement regolith development and soil shielding (Fig. \ref{fig:regolith_schematic}), which introduces a dependence on erosion rate (and therefore topographic slope) \cite{Maffre2018a}. While this introduction of regolith development into GEOCLIM is important for assessing the impact of tropical arc-continent collisions on \pCOtwo, the relatively high Ca+Mg concentration in arc rocks relative to other lithologies must also be considered.

@@

The values of four parameters within the silicate weathering component that modify the dependence of silicate weathering on temperature, runoff, and regolith thickness are poorly constrained. Rather than prescribing values, we allow these parameters and the Ca+Mg concentration of metamorphic and siliciclastic lithologies to vary within reasonable bounds (\SI; Table S2). This approach results in 93,600 unique combinations of these six parameters. For each combination, we compute spatially-resolved long-term \COtwo consumption associated with Ca+Mg fluxes using present-day runoff, temperature, and slope. We sum computed \COtwo consumption over watersheds for which data-constrained estimates are available \cite{Gaillardet1999a, Moquet2018a}, then calculate the coefficient of determination ($r^{2}$) between computed and measured \COtwo consumption in each of these watersheds. After eliminating parameter combinations that result in low $r^{2}$, 573 parameter combinations remain (\SI; Fig. S2). The resulting global \COtwo consumption of these model runs all overlap with independently-derived estimates of the global \COtwo degassing flux \cite{Gerlach2011a} as they should given that the long-term carbon cycle is in steady state (\SI; Fig. S2).

@@

Having calibrated the silicate weathering component of GEOCLIM, we use it to estimate the decrease in steady-state \pCOtwo associated with SEAI emergence. For the climate model component, we use temperature and runoff from a subset of the GFDL CM2.0 experiments \cite{Delworth2006b} (\SI). These experiments are well suited for this analysis because all non-\COtwo forcings are held constant at values representative of pre-Industrial conditions, allowing the effect of changing \pCOtwo on climatology to be isolated. Furthermore, the experiments were run long enough for the final system to approximate steady-state.

@@

\begin{SCfigure*}[\sidecaptionrelwidth][t]
    \centering
    \includegraphics[width=13.0cm]{Figures/shoreline_growth.pdf}
    \caption{The emergence of the Southeast Asian islands (also referred to as the Maritime Continent in the climate science literature) from the mid-Miocene to present. Past shorelines at 5, 10, and 15 Ma are shown in \textit{A} with associated land area summarized in \textit{B}. A significant increase in area over the past 5 million years is coincident with cooling and the onset of Northern Hemisphere glaciation as reflected in the benthic oxygen isotope record \cite{Zachos2008a} shown in \textit{C}.}
    \label{fig:shoreline_growth}
\end{SCfigure*}

To determine the position of SEAI paleoshorelines over the past 15~m.y., we use terrestrial and marine sedimentary deposits (Fig. \ref{fig:shoreline_growth}; \SI). The paleoshoreline data indicate that the Sunda-Banda Arc and New Guinea are primarily responsible for the increase in area since 15~Ma. Exhumation of the modern Sunda-Banda Arc is the result of ongoing arc-continent collision with the Australian Plate \cite{Harris2006a}. Of the major islands in this arc, only Sumatra, Belitung, Bangka, Java, Bali, and Flores were subaerially exposed before 5~Ma. However, most of Sumatra and Java along with the non-volcanic islands of the Outer Banda Arc were elevated above sea level after 5~Ma \cite{Hall2013b}. In New Guinea, emergence in the mid-Miocene is associated with collision between the Melanesian Arc and Australia's distal margin \cite{Cloos2005a}, which drove exhumation of the Irian-Marum-April Ophiolite Belt. Exhumation accelerated over the past 4~m.y. in the New Guinea Central Range due to slab-breakoff and buoyant uplift, and in eastern New Guinea due to jamming of the north-dipping subduction zone \cite{Cloos2005a}. These tectonic drivers and others throughout the region lead to progressive emergence over the past 15~m.y. that accelerated following 5~Ma (Fig. \ref{fig:shoreline_growth}B). This trend mirrors broad cooling over the Neogene that resulted in the initiation of Northern Hemisphere ice sheets (Fig. \ref{fig:shoreline_growth}C). We also include changes in areas of presently-submerged continental shelves such as the Sunda Shelf that were previously exposed (\SI; Fig. S5).

@@

Proxy-based estimates of the magnitude and trajectory of \pCOtwo change from the Miocene to the Pliocene are variable between techniques and associated assumptions underlying their interpretation \cite{Foster2017a}. The modeled \pCOtwo values for 15~Ma resemble the higher end of proxy-based \pCOtwo estimates for the early-mid-Miocene (\SI; Fig. S9), indicating that the increase in subaerially-exposed land area and tectonic topography of the SEAI is sufficient to explain long-term cooling of Earth's climate over the Neogene. The \pCOtwo threshold for Antarctic glaciation is estimated to be $\sim$750~ppm with that for Northern Hemisphere glaciation being significantly lower at $\sim$280~ppm \cite{DeConto2008a}. These modeled values of decreasing \pCOtwo associated with SEAI emergence are therefore consistent with the record of Neogene climate with Miocene ice sheets on Antarctica \cite{Sugden1995a} followed by Northern Hemisphere ice sheets developing in the Pliocene \cite{Haug2005a} as \pCOtwo subsequently decreased.

@@

Previous work has estimated that the decrease in \pCOtwo since 5~Ma associated with the emergence of the SEAI and enhanced silicate weathering is $\sim$19~ppm \cite{Molnar2015a}, in which case SEAI emergence would be a relatively minor contributor to Neogene cooling. This 19~ppm estimate was obtained using a 0-D equation that assumes a direct linear relationship between mean global temperature and changes in weathering-rate-weighted land area, scaled by a factor that is intended to account for the influence of both runoff and temperature. \pCOtwo was then estimated from the calculated temperature using a simple energy balance equation. However, the relationship between mean global temperature (or \pCOtwo) and weathering-rate-weighted land area is not linear. Furthermore, this simple linear relationship ignores spatial variability in topography and climatology, and only crudely accounts for spatial variability in lithology. In fact, the 19~ppm estimate is closer in magnitude to the decrease in \pCOtwo that we estimate if mafic volcanics associated with the Deccan Traps (a relatively flat area outside of the warm and wet tropics) are replaced with the Ca+Mg concentration of bulk continental crust (22--70 ppm; Fig. \ref{fig:scenario_pCO2}). The significant difference in steady-state \pCOtwo estimated between the ``removed Deccan Traps'' scenario and the ``paleo-SEAI'' scenarios (Fig. \ref{fig:scenario_pCO2}) demonstrates that considering changes in the spatial distribution of lithologies alone is not adequate for estimating changes in steady-state \pCOtwo. Instead, spatially-varying topography and climatology significantly modulates silicate weathering rates, and must be accounted for when estimating \pCOtwo change associated with paleogeographic change.

\section*{Alternative Mechanisms for Neogene Cooling}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=1\linewidth]{Manuscript/Figures/weatherability_curves.pdf}
    \caption{Weatherability curves for the modern and ``paleo-SEAI'' scenarios shown in Figure \ref{fig:scenario_pCO2}. Details on how these curves were generated are described in \MM. Each of the 4 curves represent a different tectonic boundary condition (i.e. the reconstructed paleoshorelines of the SEAI; Fig. \ref{fig:shoreline_growth}A) and therefore a different global weatherability. The curves show the resulting \pCOtwo for a given volcanic degassing flux such that the input flux is balanced by the silicate weathering output flux. Point B represents the pre-Industrial, in which \pCOtwo is 286~ppm. The arrow from Point A$_{1}$ to B represents the ``change in weatherability only'' scenario, in which global weatherability changes as the SEAI emerge, but the volcanic degassing flux does not change over the past 15~m.y. In this scenario, the \pCOtwo decreases from the value dictated by the 15~Ma ``paleo-SEAI'' weatherability curve ($\sim$570~ppm). The arrow from Point A$_{2}$ to B instead represents the ``change in degassing only'' scenario, in which global weatherability remains the same as the pre-Industrial, but the same change in \pCOtwo as the ``change in weatherability only'' scenario is achieved by decreasing the volcanic degassing flux from a value $\sim$13\% greater than the pre-Industrial. The arrow from Point A$_{3}$ to B represents the ``change in weatherability and degassing'' scenario, in which the same change in \pCOtwo as the other two scenarios is achieved by increasing global weatherability from our 5~Ma tectonic boundary condition and decreasing the volcanic degassing flux from a value $\sim$7\% greater than the pre-Industrial.}
    \label{fig:weatherability_curves}
\end{figure}

@@

Our model framework provides an opportunity to estimate the decrease in volcanic degassing flux necessary to achieve the same change in \pCOtwo predicted for the increase in global weatherability associated with the emergence of the SEAI over the past 15~m.y. If we use the parameter combination that had the highest $r^{2}$ between computed and measured \COtwo consumption in watersheds around the world during calibration (\textit{Calibration} and \SI; Fig. S3), GEOCLIM estimates a pre-Industrial volcanic degassing flux of 4.1$\times$10$^{12}$~mol/yr to balance the silicate weathering flux at 286~ppm. If we then assume that this volcanic degassing flux did not change over the past 15~m.y., then GEOCLIM estimates that the increase in global weatherability associated with the emergence of the SEAI lead to a change in \pCOtwo of $\sim$280~ppm (``change in weatherability only'' scenario in Fig. \ref{fig:weatherability_curves}). If we instead assume that global weatherability did not change over the past 15~m.y., then we estimate that the volcanic degassing flux needs to have been $\sim$13\% greater at 15~Ma relative to the pre-Industrial to drive the same $\sim$280~ppm change in \pCOtwo (``change in degassing only'' scenario in Fig. \ref{fig:weatherability_curves}).

This $\sim$13\% value is higher than 10\%, the highest current estimate for the volcanic degassing flux at 15~Ma relative to the present day \cite{Van-Der-Meer2014a}, suggesting that reasonable changes in the volcanic degassing flux alone are insufficient to drive the majority of cooling observed over the Neogene. However, changes in the volcanic degassing flux would have modulated changes in \pCOtwo associated with changes in global weatherability. For example, take a scenario where the increase in global weatherability associated with SEAI emergence over the past 15~m.y. is an overestimate and the increase is more accurately reflected by the difference in global weatherability between our 5~Ma and modern tectonic boundary conditions (``change in weatherability and degassing'' scenario in Fig. \ref{fig:weatherability_curves}). In this scenario, if the volcanic degassing flux decreased from a value $\sim$7\% greater than the pre-Industrial in conjunction with the smaller change in weatherability, we can achieve the same $\sim$280~ppm change in \pCOtwo.

@@

Himalayan uplift \cite{Raymo1992a, Galy2007a} has also been proposed as a driver for Neogene cooling. Rising marine \SrSr has been associated with increasing weathering of the high \SrSr Himalayan lithologies \cite{Raymo1992a}. This increase in \SrSr since ca. 35~Ma is followed by a decrease in slope ca. 15~Ma that has been attributed to exhumation of relatively low \SrSr and high \OsOs Lesser Himalayan lithologies \cite{Quade1997a, Colleps2018a}, but could also be at least partially driven by the exhumation of mafic and organic-rich forearc sediments in the SEAI during arc-continent collision. Enhanced organic matter burial in the Bay of Bengal due to high sedimentation rates could also have contributed to Neogene cooling \cite{Galy2007a}.

@@

Coupled geological constraints and modeling experiments demonstrate that the SEAI has been a growing hot spot for carbon sequestration due to silicate weathering from the Miocene to present. Changes in volcanic degassing and paleogeography elsewhere on Earth, particularly in the Himalaya and Central America, would have also effected geological carbon sources and sinks. Such changes are likely to have induced complex feedbacks on many Earth surface processes. Yet, not only does the history of SEAI emergence coincide with Neogene cooling and the onset of Northern Hemisphere glaciation, but our coupled weathering-climate model also indicates that the associated steady-state \pCOtwo change is sufficient to explain much of this cooling. These results highlight that the Earth's climate state is particularly sensitive to changes in tropical paleogeography.

@@

To create the 15~Ma ``paleo-SEAI'' curve shown in Figure \ref{fig:weatherability_curves}, we use the reconstructed paleoshorelines of the SEAI at 15~Ma (Fig. \ref{fig:shoreline_growth}A). We then select the parameter combination that had the highest $r^{2}$ between computed and measured \COtwo consumption in watersheds around the world during calibration (\textit{Calibration} and \SI; Fig. S3), and fix \pCOtwo at the 3 \pCOtwo levels at which the GFDL CM2.0 climate model experiments were computed (\SI). We then run GEOCLIM at each of these \pCOtwo levels until steady state is achieved (i.e. until the volcanic degassing flux is equal to the silicate weathering flux). We then repeat this process for the 10 and 5~Ma ``paleo-SEAI'' paleoshorelines and the present day shorelines to generate the 3 other weatherability curves. Each estimated \pCOtwo in Figure \ref{fig:scenario_pCO2} is the result of underlying weatherability curves that subtly change with the different chemical weathering parameters.

@@

\acknow{Collaborative research between N.L.S.-H. and Y.G. was initially supported by a grant from the France-Berkeley Fund. N.L.S.-H. and F.A.M. gratefully acknowledge support through NSF FRES grants \#1926001 and \#1925990. We thank Alec Brenner, Sam Lo Bianco, Mariana Lin, and Judy Pu for their data compilation contributions to the paleoshoreline reconstructions.}
